Starcode: sequence clustering based on all-pairs search

### 2014-11-07 doc/document.tex
Manuscript ready for resubmission.

We show the basic performance and scalability of starcode on a
dataset of pseudo random sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. In each
test, only one parameter is modified while the others are kept constant.

@@

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (PRJEB7686) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & 16     & single     \\
    PRJEB7686 & 127,675,537  & 50     & single     \\
    SRR950477 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

The running times of starcode, seed, slidesort, rainbow and cd-hit
are summarized in Table~\ref{tab:biobench_runtime}. We accommodated the
distance threshold for the first dataset to compensate for the reduced
sequence length. Both starcode and slidesort were executed with
the option '-d 2' and the identity for cd-hit was set to '-c 0.85'. We
were not able to run seed on dataset 1 due to limitations on the
minimum sequence length. Starcode was significantly faster than the other
tools on all the datasets. CD-HIT and seed came in second position
with a running time approximately 35 and 20 times greater on datasets
1 and 2, respectively. Rainbow was nearly an order of magnitude slower
in the job of clustering paired-end reads. We did not record the exact
running times past 10 days since this is several orders of magnitude
higher than the running time of starcode.

\begin{table}[h]
\centering
\caption{Running time (in seconds) of the software on three biological
datasets. Runs longer than 10 days were interrupted. A dash indicates
that the software cannot be used for this dataset.}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{PRJEB7686} & \textit{SRR950477} \\ \hline
    starcode   & 5           & 2,898       & 44    \\
    seed       & -           & 60,374      & -     \\
    sidesort   & 4,055       & $>$ 10 days & -     \\
    rainbow    & -           & -           & 306   \\
    cd-hit-est & 170         & 512,591     & -     \\
\end{tabular}
\label{tab:biobench_runtime}
\end{table}

The memory footprint of the different tools on the same datasets is
shown in Table~\ref{tab:biobench_memory}. The values represent the
peak memory usage throughout the run on the datasets described
above. On short reads (dataset 1), starcode outperforms the other
tools taking advantage of the trie compaction. On dataset 3, starcode
had a significantly larger memory usage than rainbow. Starcode and
cd-hit used the same amount of memory on dataset 2. Both needed twice
as much memory as slidesort, which runs in advantage because it does
not need to keep the complete graph during the all-pair comparison
step to perform future clustering.

@@

@@

\section*{Acknowledgement}
We would like to thank Maria Chatzou for her precious feedback on the
preliminary version of this manuscript and Heng-Chang Chen for
performing the Drosophila TRIP experiments used for benchmarking.

\paragraph{Funding\textcolon}
The research leading to these results has received funding from the
Government of Catalonia (Dept. of Economy and Knowledge) and the Spanish
Ministry of Economy and Competitiveness (Centro de Excelencia Severo Ochoa
2013-2017' (SEV-2012-0208). P.C. fellowship is partly financed by the
Spanish Ministry of Economy and Competitiveness (State Training Subprogram:
predoctoral fellowships for the training of PhD students (FPI) 2013).

>>>>

We show the basic performance and scalability of starcode on a
dataset of pseudorandom sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. In each
test, only one parameter is modified while the others are kept constant.

@@

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (PRJEB7686) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|c c c}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & $16\pm 1$ & single     \\
    PRJEB7686 & 127,675,537  & 50        & single     \\
    SRR950477 & 2,460,226    & 100+100     & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

The running times of starcode, seed, slidesort, rainbow and cd-hit
are summarized in Table~\ref{tab:biobench_runtime}. We accommodated the
distance threshold for the first dataset to compensate for the reduced
sequence length. Both starcode and slidesort were executed with
the option `\texttt{-d 2}' and the identity for cd-hit was set to
`\texttt{-c 0.85}'. We
were not able to run seed on dataset 1 due to limitations on the
minimum sequence length. Starcode was significantly faster than the other
tools on all the datasets. Seed and cd-hit came in second position
with a running time approximately 35 and 20 times greater on datasets
1 and 2, respectively. Rainbow was nearly an order of magnitude slower
in the job of clustering paired-end reads. We did not record the exact
running times past 10 days since this is several orders of magnitude
higher than the running time of starcode.

\begin{table}[h]
\centering
\caption{Running time (in seconds) of the software on three biological
datasets. Exact running time was not recordeed past 10 days.
A dash indicates that the software cannot be used for this dataset.}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{PRJEB7686} & \textit{SRR950477} \\ \hline
    starcode   & 5           & 2,898       & 44    \\
    seed       & -           & 60,374      & -     \\
    sidesort   & 4,055       & $>$ 10 days & -     \\
    rainbow    & -           & -           & 306   \\
    cd-hit-est & 170         & 512,591     & -     \\
\end{tabular}
\label{tab:biobench_runtime}
\end{table}

The memory footprint of the different tools on the same datasets is
shown in Table~\ref{tab:biobench_memory}. The values represent the
peak memory usage throughout the run on the datasets described
above. On short reads (dataset 1), starcode outperforms the other
tools taking advantage of the trie compaction. On dataset 3, starcode
had a significantly larger memory usage than rainbow. Starcode and
cd-hit used similar amount of memory on dataset 2. Both needed twice
as much memory as slidesort, which has the advantage of not storing
the complete graph during the all-pair comparison.

@@

It is somehwat surprising that starcode is significantly faster
than competing tools on experimental datasets, whereas seed and cd-hit
are faster on artificial datasets. Starcode was developed ground up
from TRIP experimental datasets and the poucet search was selected for
giving the best empirical results. We speculate that the trie structure
benefits from the entropy deficit that is typically observed in
experimental data versus pseudorandom reads.

@@

\section*{Acknowledgement}
We would like to thank Maria Chatzou for her precious feedback on the
preliminary version of this manuscript and Heng-Chang Chen for
performing the \textit{Drosophila} TRIP experiments.

\paragraph{Funding\textcolon}
This research was funded by the Government of Catalonia (Dept. of
Economy and Knowledge) and the Spanish
Ministry of Economy and Competitiveness (Centro de Excelencia Severo Ochoa
2013-2017' (SEV-2012-0208). P.C. fellowship is partly financed by the
Spanish Ministry of Economy and Competitiveness (State Training Subprogram:
predoctoral fellowships for the training of PhD students (FPI) 2013).
